---
title: "gutenbergr: Search and download public domain texts from Project Gutenberg"
author: "David Robinson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gutenbergr: Search and download public domain texts from Project Gutenberg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The gutenbergr package helps you download and process public domain works from the [Project Gutenberg](https://www.gutenberg.org/) collection. This includes both tools for downloading books (and stripping header/footer information), and a complete dataset of Project Gutenberg metadata that can be used to find words of interest.

### Downloading books by ID

The function `gutenberg_download()` downloads one or more works from Project Gutenberg based on their ID. For example, the book "Frankenstein; or The Modern Prometheus", by Mary Wollstonecraft Shelly, has ID 84 (see [the URL here](https://www.gutenberg.org/ebooks/84)), so `gutenberg_download(84)` downloads this text.

```{r frankenstein}
library(dplyr)
library(gutenbergr)

frankenstein <- gutenberg_download(84)

frankenstein
```

Notice it is returned as a tbl_df (a type of data frame), including two variables: `gutenberg_id` (useful if multiple books are returned), and a character vector of the text, one row per line. Notice that the header and footer added by Project Gutenberg (visible [here](http://www.gutenberg.org/files/84/84.txt)) have been stripped away.

Provide a vector of IDs to download multiple books. For example, to download Dracula (book [345](https://www.gutenberg.org/ebooks/345)) along with Frankenstein, do:

```{r frankenstein_dracula}
frankenstein_dracula <- gutenberg_download(c(84, 345), meta_fields = "title")

frankenstein_dracula
```

Notice that the `meta_fields` argument allows us to add one or more additional fields from the `gutenberg_metadata` to the downloaded text, such as title or author.

```{r}
frankenstein_dracula %>%
  count(title)
```

### Project Gutenberg Metadata

This package contains metadata for all Project Gutenberg works as R datasets, so that you can search and filter for particular works before downloading.

The dataset `gutenberg_metadata` contains information about each work, pairing Gutenberg ID with title, author, language, etc:

```{r}
gutenberg_metadata
```

For example, you could find the Gutenberg ID of Wuthering Heights by doing:

```{r}
gutenberg_metadata %>%
  filter(title == "Wuthering Heights")
```

In many analyses, you may want to filter just for English works, avoid duplicates, and include only books that have text that can be downloaded. The `gutenberg_works()` function does this pre-filtering:

```{r}
gutenberg_works()
```

It also allows you to perform filtering as an argument:

```{r}
gutenberg_works(author == "Austen, Jane")
```

#### Other meta-datasets

`gutenberg_subjects` contains pairings of works with Library of Congress subjects and topics. "lcc" means [Library of Congress Classification](https://www.loc.gov/catdir/cpso/lcco/), while "lcsh" means [Library of Congress subject headings](http://id.loc.gov/authorities/subjects.html):

```{r}
gutenberg_subjects
```

This is useful for extracting texts from a particular topic or genre, such as horror, or a particular character, such as Sherlock Holmes. The `gutenberg_id` column can then be used to download these texts or to link with other metadata.

```{r}
gutenberg_subjects %>%
  filter(subject == "Horror tales")

gutenberg_subjects %>%
  filter(grepl("Holmes, Sherlock", subject))
```

`gutenberg_authors` contains information about each author, such as aliases and birth/death year:

```{r}
gutenberg_authors
```

### Analysis

What's next after retrieving a book's text? Well, having the book as a data frame is especially useful for working with the [tidytext](https://github.com/juliasilge/tidytext) package for text analysis.

```{r}
library(tidytext)

words <- frankenstein_dracula %>%
  unnest_tokens(word, text)

words

word_counts <- words %>%
  anti_join(stop_words, by = "word") %>%
  count(title, word, sort = TRUE)

word_counts
```
